---
title: "Second Opinion: An Independent R Check of the Marketing Mix Model"
subtitle: "Does a simpler model, built separately in R, reach the same budget calls?"
author: "Lumenvale Jewelers · Growth Analytics (fictional)"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 4.5
    fig-dpi: 144
brand: _brand.yml
execute:
  warning: false
  message: false
---

::: {.callout-warning}
**This project contains synthetic data and analysis created for demonstration purposes only.**
Lumenvale Jewelers is a fictional company. Every channel, budget and result below was generated by AI.
:::

The budget recommendations in the **Channel Investment Planner** come from a Bayesian
marketing mix model written in Python (PyMC). Before the growth team moves real money,
a second analyst checks the main calls with a different, simpler method in R, reading
**the same Snowflake tables**.

The R check uses ordinary least squares with a grid search over carryover and saturation.
It has no priors and shares no code with the Python model. If the two agree on which
channels to grow and which to trim, the recommendation holds up.

```{r}
#| label: setup
source("R/lumenvale.R")
library(tidyr)
library(gt)

# To run against Snowflake (Posit Workbench managed credentials), the helper does:
# con <- DBI::dbConnect(
#   odbc::snowflake(),
#   account = "duloftf-posit-software-pbc-dev",
#   warehouse = "DEFAULT_WH",
#   database = "LUMENVALE_MMM",
#   schema = "PUBLIC"
# )
# weekly <- DBI::dbGetQuery(con, "SELECT * FROM LUMENVALE_MMM.PUBLIC.SALES_WEEKLY")
# py_summary <- DBI::dbGetQuery(con, "SELECT * FROM LUMENVALE_MMM.PUBLIC.MMM_CHANNEL_SUMMARY")

weekly <- lv_model_inputs()
py_summary <- lv_read("MMM_CHANNEL_SUMMARY")
# The generator's answer key (synthetic data only), to grade both models.
truth <- lv_read("SYNTHETIC_TRUE_PARAMETERS") |> split(~channel)
b <- lv_brand()
```

## The R model

For each channel, spend is carried over with geometric **adstock** (decay `d`) and
flattened with an **exponential saturation** `1 - exp(-adstock / s)`. Revenue is then
regressed on all eight transformed channels, plus trend, seasonality (Fourier terms and
a December-peak indicator), promotions, holidays and the two market indices. The decay
and saturation scale for each channel are chosen by coordinate descent over a grid,
minimising the residual error.

One piece of business knowledge goes in: direct-response channels (Paid Search, Email / CRM,
Affiliate) are limited to a carryover of 0.3 or less. Without that, a smooth
long-carryover series can soak up slow trends that have nothing to do with marketing.

```{r}
#| label: fit
MAX_LAG <- 13
RUN_RATE_WEEKS <- 52
DIRECT_RESPONSE <- c("Paid Search", "Email / CRM", "Affiliate")
decay_grid <- function(ch) seq(0, if (ch %in% DIRECT_RESPONSE) 0.3 else 0.9, by = 0.1)
SCALE_MULT <- c(0.5, 1, 2, 4, 8)

adstock <- function(x, decay) {
  w <- decay^(0:(MAX_LAG - 1))
  w <- w / sum(w)                             # normalised: steady spend x -> adstock x
  padded <- c(rep(mean(x[1:MAX_LAG]), MAX_LAG - 1), x)
  as.numeric(stats::filter(padded, w, sides = 1))[-(1:(MAX_LAG - 1))]
}

n <- nrow(weekly)
t <- seq_len(n)
doy <- as.numeric(format(weekly$week_start, "%j"))
fourier <- do.call(cbind, lapply(1:6, \(k) cbind(sin(2 * pi * k * doy / 365.25),
                                                  cos(2 * pi * k * doy / 365.25))))
colnames(fourier) <- paste0("f", seq_len(ncol(fourier)))
dec_peak <- as.numeric(format(weekly$week_start, "%m") == "12" &
                         as.numeric(format(weekly$week_start, "%d")) <= 24)
base_x <- cbind(trend = t / n, fourier, dec_peak,
                promo = weekly$promo_flag, holiday = weekly$holiday_flag,
                cci = as.numeric(scale(weekly$consumer_confidence_index)),
                dpi = as.numeric(scale(weekly$diamond_price_index)))

mean_spend <- sapply(LV_CHANNELS, \(ch) mean(weekly[[ch]]))
params <- tibble(channel = LV_CHANNELS, decay = 0.2, s = 2 * mean_spend)

media_matrix <- function(params) {
  sapply(seq_len(nrow(params)), \(i) {
    1 - exp(-adstock(weekly[[params$channel[i]]], params$decay[i]) / params$s[i])
  }) |> `colnames<-`(make.names(params$channel))
}
fit_lm <- function(params) lm(weekly$revenue ~ media_matrix(params) + base_x)
sse <- function(params) sum(residuals(fit_lm(params))^2)

for (pass in 1:3) {
  for (i in seq_len(nrow(params))) {
    grid <- expand_grid(decay = decay_grid(params$channel[i]), s = mean_spend[i] * SCALE_MULT)
    scores <- purrr::pmap_dbl(grid, \(decay, s) { p <- params; p$decay[i] <- decay; p$s[i] <- s; sse(p) })
    params[i, c("decay", "s")] <- grid[which.min(scores), ]
  }
}
fit <- fit_lm(params)
r2 <- summary(fit)$r.squared
mape <- mean(abs(residuals(fit)) / weekly$revenue)
```

The R model explains **`r scales::percent(r2, 0.1)`** of weekly revenue variation, with an
average error of **`r scales::percent(mape, 0.1)`**.

```{r}
#| label: rois
coefs <- coef(summary(fit))
media_rows <- paste0("media_matrix(params)", make.names(LV_CHANNELS))
X <- media_matrix(params)
recent <- (n - RUN_RATE_WEEKS + 1):n

r_results <- params |>
  mutate(
    beta = coefs[media_rows, "Estimate"],
    beta_se = coefs[media_rows, "Std. Error"],
    total_spend = sapply(channel, \(ch) sum(weekly[[ch]])),
    run_rate = sapply(channel, \(ch) mean(weekly[[ch]][recent])),
    driver = colSums(X),
    roi = beta * driver / total_spend,
    roi_se = beta_se * driver / total_spend,
    # d/dx [beta * (1 - exp(-x / s))] at the steady-state run rate
    marginal_roi = beta / s * exp(-run_rate / s),
    marginal_roi_se = beta_se / s * exp(-run_rate / s),
    negative_effect = beta < 0
  )

# The optimiser moves money from low to high marginal ROI, so the call is relative:
# the top three channels are candidates to grow, the bottom three to trim.
verdict <- function(m) {
  r <- rank(-m)
  case_when(r <= 3 ~ "Grow", r > length(m) - 3 ~ "Trim", .default = "Hold")
}

compare <- r_results |>
  select(channel, r_decay = decay, r_roi = roi, r_roi_se = roi_se,
         r_mroi = marginal_roi, r_mroi_se = marginal_roi_se) |>
  inner_join(py_summary |>
               select(channel, py_decay = decay, py_roi = roi, py_roi_p05 = roi_p05, py_roi_p95 = roi_p95,
                      py_mroi = marginal_roi, py_mroi_p05 = marginal_roi_p05, py_mroi_p95 = marginal_roi_p95),
             by = "channel") |>
  mutate(true_mroi = if (is.null(truth)) NA_real_ else sapply(channel, \(ch) truth[[ch]]$marginal_roi_at_run_rate[[1]]),
         true_roi = if (is.null(truth)) NA_real_ else sapply(channel, \(ch) truth[[ch]]$roi[[1]]),
         r_verdict = verdict(r_mroi), py_verdict = verdict(py_mroi),
         agree = r_verdict == py_verdict,
         channel = factor(channel, levels = rev(LV_CHANNELS)))

rank_corr <- cor(compare$r_mroi, compare$py_mroi, method = "spearman")
n_agree <- sum(compare$agree)
```

## Where the next dollar should go: R vs Python

The question that drives the budget is the **marginal ROI**: what the next dollar in
each channel returns at today's spend.

```{r}
#| label: fig-mroi
#| fig-cap: "Marginal ROI at today's spend. Python Bayesian MMM (90% credible interval), independent R model (±1.645 SE), and the true value used to simulate the data. In the shaded zone below 1×, the next dollar returns less than it costs."
plot_df <- bind_rows(
  compare |> transmute(channel, model = "Python · Bayesian MMM", est = py_mroi, lo = py_mroi_p05, hi = py_mroi_p95),
  compare |> transmute(channel, model = "R · independent check", est = r_mroi,
                       lo = r_mroi - 1.645 * r_mroi_se, hi = r_mroi + 1.645 * r_mroi_se)
) |> mutate(model = factor(model, levels = c("Python · Bayesian MMM", "R · independent check")))

X_MAX <- 6  # a thinly-funded channel can have a huge, very uncertain OLS estimate
off_scale <- plot_df |> filter(est > X_MAX)
plot_df <- plot_df |> mutate(lo = pmax(lo, 0), hi = pmin(hi, X_MAX), est = pmin(est, X_MAX))

ggplot(plot_df, aes(y = channel, x = est, colour = model)) +
  annotate("rect", xmin = -Inf, xmax = 1, ymin = -Inf, ymax = Inf, fill = b[["well"]], alpha = 0.6) +
  geom_vline(xintercept = 1, colour = b[["secondary"]], linetype = "dashed", linewidth = 0.4) +
  geom_linerange(aes(xmin = lo, xmax = hi), position = position_dodge(width = 0.6), linewidth = 1) +
  geom_point(position = position_dodge(width = 0.6), size = 2.6) +
  geom_point(data = compare, aes(y = channel, x = true_mroi), inherit.aes = FALSE,
             shape = 18, size = 3.4, colour = b[["foreground"]]) +
  scale_colour_manual(values = c(b[["primary"]], b[["secondary"]])) +
  geom_text(data = off_scale, aes(y = channel, x = X_MAX, label = sprintf("R: %.1f×, off scale", est)),
            inherit.aes = FALSE, hjust = 1, nudge_y = -0.35, size = 3.2, colour = b[["secondary"]]) +
  scale_x_continuous(labels = \(x) paste0(x, "×"), limits = c(0, X_MAX), oob = scales::squish) +
  labs(x = "Revenue returned by the next $1", y = NULL,
       subtitle = "Black diamond: the true value, known only because the data is synthetic") +
  theme_lumenvale(base_size = 13)
```

```{r}
#| label: tbl-compare
compare |>
  arrange(match(channel, LV_CHANNELS)) |>
  mutate(channel = as.character(channel),
         agree = if_else(agree, "✓", "✗")) |>
  select(channel, py_roi, r_roi, py_mroi, r_mroi, true_mroi, py_verdict, r_verdict, agree) |>
  gt() |>
  tab_header("Do the two models make the same budget call?",
             "Verdict from each model's marginal ROI ranking: top three grow, bottom three trim") |>
  tab_spanner("Average ROI", c(py_roi, r_roi)) |>
  tab_spanner("Marginal ROI", c(py_mroi, r_mroi, true_mroi)) |>
  tab_spanner("Verdict", c(py_verdict, r_verdict, agree)) |>
  cols_label(channel = "Channel", py_roi = "Python", r_roi = "R", py_mroi = "Python", r_mroi = "R",
             true_mroi = "Truth", py_verdict = "Python", r_verdict = "R", agree = "Same?") |>
  fmt_number(c(py_roi, r_roi, py_mroi, r_mroi, true_mroi), decimals = 2, pattern = "{x}×") |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = agree)) |>
  tab_style(style = cell_text(color = b[["danger"]]),
            locations = cells_body(columns = agree, rows = agree == "✗")) |>
  tab_source_note("This project contains synthetic data and analysis created for demonstration purposes only.")
```

**Result:** the two models make the same call on **`r n_agree` of 8** channels, and rank
the channels by marginal ROI with a correlation of **`r round(rank_corr, 2)`**.
`r if (all(compare$r_mroi[compare$channel == "Paid Search"] < compare$r_mroi[compare$channel == "Connected TV"])) "Both say the next dollar works harder in **Connected TV** than in **Paid Search**, which is the core recommendation in the planner." else "The R model does not confirm that Connected TV beats Paid Search at the margin, so the headline recommendation needs a closer look."`

## Carryover

```{r}
#| label: fig-decay
#| fig-cap: "Estimated weekly carryover (adstock decay) by channel."
compare |>
  select(channel, Python = py_decay, R = r_decay) |>
  mutate(Truth = if (is.null(truth)) NA_real_ else sapply(as.character(channel), \(ch) truth[[ch]]$decay[[1]])) |>
  pivot_longer(-channel, names_to = "source", values_to = "decay") |>
  mutate(source = factor(source, levels = c("Python", "R", "Truth"))) |>
  ggplot(aes(y = channel, x = decay, fill = source)) +
  geom_col(position = position_dodge2(width = 0.75, reverse = TRUE), width = 0.7) +
  scale_fill_manual(values = c(Python = b[["primary"]], R = b[["secondary"]], Truth = b[["foreground"]])) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "Share of the effect carried into the next week", y = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_lumenvale(base_size = 13)
```

Both models find that **Connected TV** and **Podcast** carry over for weeks, while
**Paid Search** and **Email / CRM** act almost at once.

## Where they differ, and why that is fine

- The R model has no priors, so small channels with little spend variation swing
  widely. **Email / CRM** is the clearest case: R puts its marginal ROI far above every other
  channel, with a very wide interval. The Bayesian model shrinks it toward sensible
  values instead. Both models still rank Email / CRM near the top, but a figure like that
  should never be used on its own to set a budget.
- OLS standard errors ignore the uncertainty in the chosen decay and saturation, so the
  R intervals are too narrow. Use them to judge direction, not precision.
- Where the verdicts differ, the planner's **Business constraints** panel is the place
  to cap exposure until a test (for example a geo holdout) settles it.

```{r}
#| label: save
lv_write("MMM_R_CROSSCHECK", compare |>
           mutate(channel = as.character(channel), checked_at = as.POSIXct(format(Sys.time(), tz = "UTC"), tz = "UTC")) |>
           select(channel, r_decay, r_roi, r_mroi, py_roi, py_mroi, r_verdict, py_verdict, agree, checked_at))
```

*Data source: `r lv_source()$label`. The R results are saved as `MMM_R_CROSSCHECK` next
to the Python model outputs.*
