---
title: "Weekly Marketing Mix Readout"
subtitle: "Lumenvale Jewelers (fictional) · Growth team"
date: today
format:
  email:
    toc: false
    fig-dpi: 144
brand: _brand.yml
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-warning}
**This project contains synthetic data and analysis created for demonstration purposes only.**
Lumenvale Jewelers is a fictional company. Every channel, budget and result below was generated by AI.
:::

```{r}
#| label: setup
source("R/lumenvale.R")
library(gt)

# On Posit Workbench or Connect the helper reads Snowflake, like this:
# con <- DBI::dbConnect(
#   odbc::snowflake(),
#   account = "duloftf-posit-software-pbc-dev",
#   warehouse = "DEFAULT_WH",
#   database = "LUMENVALE_MMM",
#   schema = "PUBLIC"
# )
# sales <- DBI::dbGetQuery(con, "SELECT * FROM LUMENVALE_MMM.PUBLIC.SALES_WEEKLY")
# scenarios <- DBI::dbGetQuery(con, "SELECT * FROM LUMENVALE_MMM.PUBLIC.MMM_SAVED_SCENARIOS")

b <- lv_brand()
sales <- lv_read("SALES_WEEKLY") |> mutate(week_start = as.Date(week_start))
spend <- lv_read("MARKETING_SPEND_WEEKLY") |> mutate(week_start = as.Date(week_start))
contrib <- lv_read("MMM_CONTRIBUTIONS_WEEKLY") |> mutate(week_start = as.Date(week_start))
summary_py <- lv_read("MMM_CHANNEL_SUMMARY")
scenarios <- lv_read("MMM_SAVED_SCENARIOS")
crosscheck <- tryCatch(lv_read("MMM_R_CROSSCHECK"), error = function(e) NULL)

weekly <- sales |>
  group_by(week_start) |>
  summarise(revenue = sum(revenue), orders = sum(orders), .groups = "drop") |>
  left_join(spend |> group_by(week_start) |> summarise(spend = sum(spend), .groups = "drop"), by = "week_start") |>
  left_join(contrib |> filter(component != "Baseline") |>
              group_by(week_start) |> summarise(media = sum(contribution), .groups = "drop"), by = "week_start") |>
  arrange(week_start)

last <- tail(weekly, 1)
prev <- weekly[nrow(weekly) - 1, ]
yago <- weekly |> slice_min(abs(as.numeric(week_start - (last$week_start - 364))), n = 1, with_ties = FALSE)
chg <- function(a, b) scales::percent(a / b - 1, accuracy = 0.1, style_positive = "plus")
# Quarto's email filter cannot handle a literal percent sign, so the email says it in words.
chg_words <- function(a, b) {
  d <- a / b - 1
  if (abs(d) < 0.0005) return("flat")
  sprintf("%s %.1f percent", if (d >= 0) "up" else "down", abs(100 * d))
}
week_label <- format(last$week_start, "%b %d, %Y")
media_roas <- last$media / last$spend

# Latest version of each saved scenario, best first.
saved <- if (nrow(scenarios) > 0) {
  scenarios |>
    mutate(saved_at = as.POSIXct(saved_at, tz = "UTC")) |>
    group_by(scenario_name) |>
    filter(saved_at == max(saved_at)) |>
    summarise(saved_at = first(saved_at), saved_by = first(saved_by), budget = first(total_budget),
              revenue = first(total_revenue), roas = first(blended_roi), lift = first(change_vs_current),
              ctv = sum(quarterly_budget[channel == "Connected TV"]),
              search = sum(quarterly_budget[channel == "Paid Search"]), .groups = "drop") |>
    arrange(desc(lift))
} else tibble()
# The planner rounds today's quarterly budget to the nearest $10K; do the same.
current_budget <- function(ch) round(summary_py$current_quarter_budget[summary_py$channel == ch], -4)
current_ctv <- current_budget("Connected TV")
current_search <- current_budget("Paid Search")
best <- if (nrow(saved) > 0) saved[1, ] else NULL

saturated <- summary_py |> filter(marginal_roi < 1) |> pull(channel)
top_mroi <- summary_py |> slice_max(marginal_roi, n = 1)
```

```{r}
#| label: plots
recent <- tail(weekly, 26) |>
  mutate(baseline = revenue - media) |>
  tidyr::pivot_longer(c(baseline, media), names_to = "part", values_to = "value") |>
  mutate(part = factor(part, levels = c("baseline", "media"),
                       labels = c("Baseline demand", "Driven by media (model estimate)")))

p_revenue <- ggplot(recent, aes(week_start, value, fill = part)) +
  geom_col(width = 5, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = c(b[["well"]], b[["primary"]])) +
  scale_y_continuous(labels = \(x) lv_money(x, 0.1)) +
  scale_x_date(date_labels = "%b %d",
               breaks = \(lims) seq(as.Date(format(lims[1] + 31, "%Y-%m-01")), lims[2] - 7, by = "month")) +
  labs(x = NULL, y = NULL, title = "Weekly revenue, last 26 weeks") +
  theme_lumenvale(base_size = 13)

p_mroi <- summary_py |>
  mutate(channel = reorder(channel, marginal_roi)) |>
  ggplot(aes(marginal_roi, channel)) +
  annotate("rect", xmin = -Inf, xmax = 1, ymin = -Inf, ymax = Inf, fill = b[["well"]], alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = b[["secondary"]], linewidth = 0.4) +
  geom_linerange(aes(xmin = marginal_roi_p05, xmax = marginal_roi_p95), colour = "#A9B8C4", linewidth = 1.2) +
  geom_point(aes(colour = channel), size = 3.5, show.legend = FALSE) +
  scale_colour_manual(values = LV_CHANNEL_COLORS) +
  scale_x_continuous(labels = \(x) paste0(x, "×")) +
  labs(x = "Revenue from the next $1 at today's spend (90% interval)", y = NULL,
       title = "Where the next dollar works hardest") +
  theme_lumenvale(base_size = 13)
```

## Week of `r week_label`

```{r}
#| label: kpis
tibble(
  Metric = c("Revenue", "Orders", "Media spend", "Media-driven revenue (model)", "Media ROAS"),
  `This week` = c(lv_money(last$revenue), scales::comma(last$orders), lv_money(last$spend),
                  lv_money(last$media), sprintf("%.2f×", media_roas)),
  `vs last week` = c(chg(last$revenue, prev$revenue), chg(last$orders, prev$orders),
                     chg(last$spend, prev$spend), chg(last$media, prev$media), ""),
  `vs last year` = c(chg(last$revenue, yago$revenue), chg(last$orders, yago$orders),
                     chg(last$spend, yago$spend), chg(last$media, yago$media), "")
) |>
  gt() |>
  cols_align("right", -Metric) |>
  tab_options(table.width = pct(100))
```

```{r}
#| label: plot-revenue
#| fig-height: 3.6
p_revenue
```

## What the model says

```{r}
#| label: plot-mroi
#| fig-height: 3.8
p_mroi
```

- **`r paste(saturated, collapse = ", ")`** `r if (length(saturated) == 1) "is" else "are"` past
  the point of diminishing returns: the next dollar returns less than it costs.
- **`r top_mroi$channel`** has the most headroom: about **`r sprintf("%.1f×", top_mroi$marginal_roi)`**
  revenue from the next dollar.
`r if (!is.null(crosscheck)) sprintf("- An independent R model agrees on **%d of 8** channel calls (see the model cross-check report).", sum(crosscheck$agree))`

## Scenarios the team is considering

```{r}
#| label: scenarios
if (nrow(saved) == 0) {
  cat("No scenarios saved yet. Save one from the **Scenario** or **Plan the full mix** tab of the Channel Investment Planner.")
} else {
  saved |>
    slice_head(n = 5) |>
    transmute(Scenario = scenario_name, `Saved by` = saved_by,
              `Paid Search` = search - current_search, `Connected TV` = ctv - current_ctv,
              `Projected revenue` = revenue, `vs current plan` = lift, ROAS = roas) |>
    gt() |>
    tab_spanner("Budget change per quarter", c(`Paid Search`, `Connected TV`)) |>
    fmt_currency(c(`Paid Search`, `Connected TV`, `vs current plan`), decimals = 1, suffixing = TRUE,
                 force_sign = TRUE) |>
    fmt_currency(`Projected revenue`, decimals = 2, suffixing = TRUE) |>
    fmt_number(ROAS, decimals = 2, pattern = "{x}×") |>
    tab_style(cell_text(weight = "bold", color = b[["success"]]), cells_body(columns = `vs current plan`)) |>
    tab_options(table.width = pct(100))
}
```

::: {.email}

::: {.subject}
Lumenvale weekly mix readout · `r week_label`: revenue `r lv_money(last$revenue)` `r chg_words(last$revenue, yago$revenue)` on last year
:::

**This project contains synthetic data and analysis created for demonstration purposes only.**

Hi team,

Here is the marketing mix readout for the week of **`r week_label`**.

- Revenue was **`r lv_money(last$revenue)`**, `r chg_words(last$revenue, prev$revenue)` on last week
  and `r chg_words(last$revenue, yago$revenue)` on the same week last year.
- Media spend was **`r lv_money(last$spend)`**; the model credits media with
  **`r lv_money(last$media)`** of revenue (`r sprintf("%.2f×", media_roas)` ROAS).
- **`r paste(saturated, collapse = ", ")`** `r if (length(saturated) == 1) "is" else "are"` saturated;
  **`r top_mroi$channel`** has the most headroom (`r sprintf("%.1f×", top_mroi$marginal_roi)` from the next dollar).
`r if (!is.null(best)) sprintf("- The strongest saved scenario, **%s** (saved by %s), projects **%s** per quarter versus the current plan.", best$scenario_name, best$saved_by, scales::label_currency(accuracy = 0.01, scale_cut = scales::cut_short_scale(), style_positive = 'plus')(best$lift)) else "- No scenarios have been saved in the planner yet."`

```{r}
#| label: email-plot
#| fig-height: 3.6
#| fig-width: 7
p_revenue
```

Plan your own scenario in the **Channel Investment Planner** on Posit Connect.

*Data: `r lv_source()$label`. Synthetic data; Lumenvale Jewelers is fictional.*

:::
